---
title: "First EDA"
author: "Saverio Fontana"
date: "2025-03-16"
output: html_document
---

```{r setup, include=FALSE}
# Loading necessary libraries
library(arrow)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(DataExplorer)
knitr::opts_chunk$set(echo = TRUE)
```



```{r loading data}
# Read the Parquet file and load the data
data <- read_parquet("data/processed/set_a.parquet")
data$Time <- as.POSIXct(data$Time, tz = "UTC")  # Set time zone explicitly

# Check the first few timestamps
head(data$Time)  # Output should show the exact timestamps, e.g., "2025-03-10 01:00:00 UTC"

# Visualizing missing data
plot_missing(data)
```
## Initial Data Exploration

```{r}
# Check the dimensions of the dataset
dim(data)

# Check the column names
colnames(data)

# Summary statistics for numeric columns
summary(data)



# Histograms for numeric variables (ignoring NAs)
data %>%
  select(where(is.numeric)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Distributions of Numeric Variables (Ignoring NAs)")
```

### Variable Explanation and Notes

#### Variable Descriptions (general descriptors and time series)

"These six descriptors are collected at the time the patient is admitted to the ICU. 
Their associated time-stamps are 00:00 (thus they appear at the beginning of each 
patient's record)."

- **RecordID**: A unique integer for each ICU stay
- **Age**: Age in years
- **Gender**: Gender (0: female, 1: male)
- **Height**: Height in cm
- **ICUType**: ICU type (1: Coronary Care Unit, 2: Cardiac Surgery Recovery Unit, 3: Medical ICU, or 4: Surgical ICU)
- **Weight**: Weight in kg

"NB: All valid values for general descriptors, time series variables, and outcome-related 
descriptors are non-negative (≥ 0). A value of -1 indicates missing or unknown data."



```{r}
#So first of all we set the -1 in columns "Age", "Gender", "Height", ICUType, Weight to NA
# Replace -1 with NA for Height, Age, and Weight
data <- data %>%
  mutate(
    Height = ifelse(Height == -1, NA, Height),
    Age = ifelse(Age == -1, NA, Age),
    Weight = ifelse(Weight == -1, NA, Weight),
    ICUType = ifelse(ICUType == -1, NA, ICUType),
    Gender = ifelse(Gender ==-1, NA, Gender)
  )


#So when we will use algoritms that need to remove the NAs? What will we do? 
#I mean we could even do imputation actually (Weight can be predicted by age and Height for ex)
```


###Handling missing data in static variables
```{r}
# Static data: Keep the first row for each patient for simplicity in the next EDA
#and inspect for missing values
data_static <- data %>% group_by(RecordID) %>% slice(1)

# Frequency table for categorical variables (Gender, ICUType)
table(data_static$Gender, useNA = "ifany") 
table(data_static$ICUType, useNA = "ifany")

# Bar plots for categorical variables (Gender and ICUType)
data_static %>%
  ggplot(aes(x = factor(Gender))) +
  geom_bar() +
  labs(title = "Gender Distribution")

data_static %>%
  ggplot(aes(x = factor(ICUType))) +
  geom_bar() +
  labs(title = "ICU Type Distribution (Excluding NAs)")

```




### Exploring Continuous Static Variables (Height, Weight, Age)

```{r}
# Distribution for Height
data_static %>%
  filter(!is.na(Height)) %>%
  ggplot(aes(x = Height)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") + 
  labs(title = "Distribution of Height", x = "Height", y = "Count")


# Density plot for Height
data_static %>%
  filter(!is.na(Height)) %>%
  ggplot(aes(x = Height)) +
  geom_density(fill = "skyblue", alpha = 0.5) +  
  labs(title = "Density Plot of Height", x = "Height", y = "Density")

# Inspect unusually low or high Height values
outlier_data <- data_static %>%
  filter((Height < 100 | Height > 300) & Height != -1) %>%
  select(RecordID, Age, Height, Weight)

#upon inspection I'd suggest to set their values of Height to NA

# Extract the RecordID from the filtered data
outlier_ids <- outlier_data$RecordID

# Set Height to NA for the identified records in the original data
data_clean <- data %>%
  mutate(Height = ifelse(RecordID %in% outlier_ids, NA, Height))

# Distribution for Weight
data_static %>%
  filter(Weight != -1) %>%
  ggplot(aes(x = Weight)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") + 
  labs(title = "Distribution of Weight", x = "Weight", y = "Count")

# Distribution for Age (NB, no values over 90, so maybe data are right truncated)
data_static %>%
  filter(Age != -1) %>%
  ggplot(aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") + 
  labs(title = "Distribution of Age", x = "Age", y = "Count")

# Smooth density plot for Weight
data_static %>%
  filter(Weight != -1) %>%
  ggplot(aes(x = Weight)) +
  geom_density(fill = "skyblue", alpha = 0.5) +  
  labs(title = "Density Plot of Weight", x = "Weight", y = "Density")

# Smooth density plot for Age
data_static %>%
  filter(Age != -1) %>%
  ggplot(aes(x = Age)) +
  geom_density(fill = "skyblue", alpha = 0.5) +  
  labs(title = "Density Plot of Age", x = "Age", y = "Density")

###########
# Inspecting Extreme Weight Values
###########

# Inspect unusually low or high Weight values
data_static %>%
  filter((Weight < 30 | Weight > 300) & Weight != -1) %>%  
  select(RecordID, Age, Height, Weight)
# Don't know actually, maybe it's an ok value honestly


#put also the summaries in the output
```
## Boxplot for Age by ICUType

```{r}
data_static %>%
  filter(Age != -1) %>%
  ggplot(aes(x = factor(ICUType), y = Age, fill = factor(ICUType))) +
  geom_boxplot() +
  labs(title = "Age by ICU Type", x = "ICU Type", y = "Age") +
  theme_minimal()

# Boxplot for Weight by ICUType
data_static %>%
  filter(Weight != -1) %>%
  ggplot(aes(x = factor(ICUType), y = Weight, fill = factor(ICUType))) +
  geom_boxplot() +
  labs(title = "Weight by ICU Type", x = "ICU Type", y = "Weight") +
  theme_minimal()

# Boxplot for Height by ICUType
data_static %>%
  filter(Height != -1) %>%
  ggplot(aes(x = factor(ICUType), y = Height, fill = factor(ICUType))) +
  geom_boxplot() +
  labs(title = "Height by ICU Type", x = "ICU Type", y = "Height") +
  theme_minimal()
```

NOTE, I also tried to do some anova to see if the difference between ICU class is signficant. 
I get non signficant results for everything exept for Age, but then since the Age distribution is soo far from a Normal the results are not reliable.
