---
title: "First EDA + Data Cleaning"
author: "Saverio Fontana"
date: "2025-03-16"
output: html_document
---

```{r setup, include=FALSE}
# Loading necessary libraries
library(arrow)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(DataExplorer)
library(here)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())

```

# Data loading
```{r}
# Read the Parquet file and load the data
data <- read_parquet("data/processed/set_a/set_a.parquet")

data$Time <- as.POSIXct(data$Time, tz = "UTC")  # Set time zone explicitly

# Check the first few timestamps
head(data$Time)  # Output should show the exact timestamps, e.g., "2025-03-10 01:00:00 UTC"

# Visualizing missing data (a lot)
plot_missing(data)
```
## Initial Data Exploration

```{r}
# Check the dimensions of the dataset
dim(data)

# Check the column names
colnames(data)

# Summary statistics for numeric columns
summary(data)
```

### Variable Explanation and Notes

#### Variable Descriptions (general descriptors and time series)

"These six descriptors are collected at the time the patient is admitted to the ICU. 
Their associated time-stamps are 00:00 (thus they appear at the beginning of each 
patient's record)."

- **RecordID**: A unique integer for each ICU stay
- **Age**: Age in years
- **Gender**: Gender (0: female, 1: male)
- **Height**: Height in cm
- **ICUType**: ICU type (1: Coronary Care Unit, 2: Cardiac Surgery Recovery Unit, 3: Medical ICU, or 4: Surgical ICU)
- **Weight**: Weight in kg

"*NB**: All valid values for general descriptors, time series variables, and outcome-related 
descriptors are non-negative (≥ 0). A value of -1 indicates missing or unknown data."


# Data Handling and extreme values
```{r}
#So first of all we set the -1 in columns "Age", "Gender", "Height", ICUType, Weight to NA
# Replace -1 with NA for Height, Age, and Weight
data <- data %>%
  mutate(
    Height = ifelse(Height == -1, NA, Height),
    Age = ifelse(Age == -1, NA, Age),
    Weight = ifelse(Weight == -1, NA, Weight),
    ICUType = ifelse(ICUType == -1, NA, ICUType),
    Gender = ifelse(Gender ==-1, NA, Gender)
  )

```


## Handling missing data in static variables

```{r}
# Static data: Keep the first row for each patient for simplicity in the next EDA
#and inspect for missing values
data_static <- data %>% group_by(RecordID) %>% slice(1)

# Frequency table for categorical variables (Gender, ICUType)
table(data_static$Gender, useNA = "ifany") 
table(data_static$ICUType, useNA = "ifany")

# Bar plots for categorical variables (Gender and ICUType)
data_static %>%
  ggplot(aes(x = factor(Gender))) +
  geom_bar() +
  labs(title = "Gender Distribution")

data_static %>%
  ggplot(aes(x = factor(ICUType))) +
  geom_bar() +
  labs(title = "ICU Type Distribution (Excluding NAs)")

```




## Exploring Continuous Static Variables (Height, Weight, Age)

```{r}
# Distribution for Height
data_static %>%
  filter(!is.na(Height)) %>%
  ggplot(aes(x = Height)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") + 
  labs(title = "Distribution of Height", x = "Height", y = "Count")


# Density plot for Height
data_static %>%
  filter(!is.na(Height)) %>%
  ggplot(aes(x = Height)) +
  geom_density(fill = "skyblue", alpha = 0.5) +  
  labs(title = "Density Plot of Height", x = "Height", y = "Density")

# Inspect unusually low or high Height values
outlier_data <- data_static %>%
  filter((Height < 100 | Height > 300) & Height != -1) %>%
  select(RecordID, Age, Height, Weight)

#upon inspection I'd suggest to set their values of Height to NA (maybe 1.8 for example can mean 180 cm but since we are not sure it's better to be conservative)

# Extract the RecordID from the filtered data
outlier_ids <- outlier_data$RecordID

# Set Height to NA for the identified records in the original data
data_clean <- data %>%
  mutate(Height = ifelse(RecordID %in% outlier_ids, NA, Height))

# Distribution for Weight
data_static %>%
  filter(Weight != -1) %>%
  ggplot(aes(x = Weight)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") + 
  labs(title = "Distribution of Weight", x = "Weight", y = "Count")

# Distribution for Age (NB, no values over 90, so maybe data are right truncated)
data_static %>%
  filter(Age != -1) %>%
  ggplot(aes(x = Age)) +
  geom_histogram(binwidth = 5, 
                 fill = "#4E79A7",           # More professional blue
                 color = "white",             # White borders for cleaner look
                 alpha = 0.8) +               # Slightly transparent
  labs(title = "Patient Age Distribution",    # More descriptive title
       subtitle = "Ages binned in 5-year intervals",
       x = "Age (years)", 
       y = "Number of Patients") +            # More clinical-sounding label
  theme_minimal() +                           # Clean, modern background
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5, color = "gray40"),
    axis.title = element_text(size = 12),
    panel.grid.major = element_line(color = "gray90"),  # Lighter grid lines
    panel.grid.minor = element_blank()                  # Remove minor grid
  ) +
  scale_x_continuous(breaks = seq(0, 120, by = 10))     # Better x-axis labels



###########
# Inspecting Extreme Weight Values
###########

# Inspect unusually low or high Weight values
data_static %>%
  filter((Weight < 30 | Weight > 300) & Weight != -1) %>%  
  select(RecordID, Age, Height, Weight)
# Don't know actually, maybe it's an ok value honestly. Super skinny old lady


#put also the summaries in the output
```


# Exploring Time changing variables

This step is performed to inspect the variable distribution and choose for each one of them the correct scaling method 
```{r}


```
```{r}

#Variables handling 

summary(data$Albumin) #(g/dL)
hist(data$Albumin)
"Albumin ok
bell shaped"

summary(data$ALP) #(IU/L)]
hist(data$ALP) #left skewed
#usually max range is around 150.
#Maybe should remove over 1000, 


summary(data$ALT) #(IU/L)]
hist(data$ALT) #left skewed
#maybe should remove over 3000, 


summary(data$AST) #(IU/L)]
hist(data$AST) #left skewed
##maybe should remove over 3000, 

summary(data$Bilirubin) #(mg/dL)
hist(data$Bilirubin) #left skewed


summary(data$BUN) #(mg/dL)
hist(data$BUN) #left skewed
#GOOD VALUES


summary(data$Cholesterol) #(mg/dL)
hist(data$Cholesterol) #bell shaped
#okish values


summary(data$Creatinine) #(mg/dL)
hist(data$Creatinine) #left skewed
#okish 


summary(data$DiasABP) #(mmHg)
hist(data$DiasABP) #normali-ish
"Set to NaN if < 1	Unrealistic diastolic pressure
Set to NaN if > 200	Very high diastolic pressure, likely error"

summary(data$FiO2) #(0-1)
hist(data$FiO2) #impossible to define 


summary(data$GCS) #Glasgow Coma Scale (3-15) 
hist(data$GCS) #right skewed, makes sense, they are in coma ahahha

summary(data$Glucose) #(mg/dL)
hist(data$Glucose) #a bit normal (left skewed)

#find the columns RecordID with values over 1000
high_glucose_records <- data %>%
  filter(Glucose > 1000)

# Display the RecordID(s) with Glucose > 1000
IDs <- high_glucose_records$RecordID

#LET'S INSPECT One of them 
Unusual_patient <- data[data$RecordID == IDs[1], ]
Unusual_patient$Glucose
#An incredible drop in one hour. Don't know wheter we should cancel the over 1000



summary(data$HCO3) #(mmol/L)
hist(data$HCO3) #normal-ish


summary(data$HCT) #(%)
hist(data$HCT) #normal-ish

summary(data$HR) #(bpm)
hist(data$HR) #normal-ish
#There are some zeros, inspect:
high_HR_records <- data %>%
  filter(HR == 0)

#inspect one of them 
Unusual_patient <- data[data$RecordID == high_HR_records$RecordID[1], ]
Unusual_patient
#here, the 0 is most likely a mistake, or maybe a collapse and then a revitalization? 



Unusual_patient <- data[data$RecordID == high_HR_records$RecordID[3], ]
Unusual_patient
#but here ffor example no, cause it's the last recorded value and probably he is dead...



summary(data$K) #(mEq/L)
hist(data$K) #normal-ish-ish


summary(data$Lactate) #(mmol/L)
hist(data$Lactate) #left skewed
#some extreme values... 


summary(data$Mg) #(mmol/L)
hist(data$Mg) #normal-ish w some extreme values 


summary(data$MAP)
hist(data$MAP) #normal-ish
"MAP	Set to NaN if < 1	Impossible mean arterial pressure
Over 200?"

summary(data$MechVent) #1 if yes.... PROBLEM!!
#We only have ones --> this mean that the others are NAN, here we need to choose wether NAN means 0
#I suggest to put a 1 to patients who have alredy a 1, and put 0 otwise.
data$MechVent <- as.numeric(as.character(data$MechVent))

data <- data %>%
  group_by(RecordID) %>%
  mutate(MechVent = ifelse(any(MechVent == 1, na.rm = TRUE), 
                         replace(MechVent, is.na(MechVent), 1),
                         replace(MechVent, is.na(MechVent), 0))) %>%
  ungroup()

summary(data$MechVent)

#change again to categorical
data$MechVent <- as.factor(data$MechVent)
table(data$MechVent)



summary(data$Na) #(mEq/L)
hist(data$Na) #normal-ish



summary(data$NIDiasABP) #(mmHg)
hist(data$NIDiasABP) #normal-ish


summary(data$NIMAP) #(mmHg)
hist(data$NIMAP) #normal-ish

summary(data$NISysABP) #(mmHg)
hist(data$NISysABP) #normal-ish





summary(data$SysABP) #(mmHg)
hist(data$SysABP) #normal-ish
"Set to NaN if < 1	Impossible systolic pressure
Over 200?"

summary(data$PaCO2) #(mmHg)
hist(data$PaCO2) #normal-ish, but the right tail is really big



summary(data$PaO2) #(mmHg)
hist(data$PaO2) #normal-ish, but the right tail is really big 
#inspect values smaller than 1
low_PaO2_records <- data %>%
  filter(PaO2 < 1)

# Display the RecordID(s) with PaO2 < 1
IDs <- low_PaO2_records$RecordID

#LET'S INSPECT One of them
Unusual_patient <- data[data$RecordID == IDs[1], ]
Unusual_patient
#the 0 seems more a missing value. 



#smaller than 20?
low_PaO2_records <- data %>%
  filter(PaO2 < 20)

# Display the RecordID(s) with PaO2 < 20
IDs <- low_PaO2_records$RecordID

#One of them is the one with 0, let's see the other
Unusual_patient <- data[data$RecordID == IDs[2], ]
Unusual_patient
#it's more likely to be 74.7


#Correct the non missing data changing 0 to Nan and changining the 74.7
data$PaO2[data$PaO2 == 0] <- NA
data$PaO2[data$PaO2 == 7.47] <- 74.7



summary(data$pH) #pH
# Multiply by 0.1 if between 65 and 80 (likely unit error)
idx <- which(!is.na(data$pH) & data$pH >= 65 & data$pH <= 80)
data$pH[idx] <- data$pH[idx] * 0.1

# Multiply by 0.01 if between 650 and 800 (likely unit error)
idx <- which(!is.na(data$pH) & data$pH >= 650 & data$pH <= 800)
data$pH[idx] <- data$pH[idx] * 0.01

hist(data$pH) #normal-ish


summary(data$Platelets) #(cells/nL)
hist(data$Platelets) #normal-ish ( a bit left skewed)



summary(data$RespRate) #(bpm)
hist(data$RespRate) #normal-ish

#same as HR, maybe mistakes, maybe dead, maybe ventilation
 



summary(data$SaO2) #(%)
hist(data$SaO2) #right skewed


summary(data$Temp) #(°C)
#probably mistakes in recording, also cause there are 17 patients w/ a value of -17.8 which we proceed to remove

data$Temp[!is.na(data$Temp) & data$Temp < 20] <- NA #under 20 is likely incompatible with life


summary(data$Temp)
hist(data$Temp) #normal-ish


summary(data$TroponinI) #(ng/mL)
hist(data$TroponinI) #left skewed
#some extreme values


summary(data$TroponinT) #(ng/mL)
hist(data$TroponinT) #left skewed
#some extreme values

summary(data$Urine) #(mL)
hist(data$Urine) #left skewed
#Some really extreme values

summary(data$WBC) #(cells/nL)
hist(data$WBC) #left skewed
#some extreme values


```
## Final Steps to ensure consistency of data before coding everything in Python

```{r}
#find if there are any RecordID like this:140501, 150649, 140936, 143656, 141264, 145611, 142998, 147514, 142731,150309, 155655, 156254.
data %>%
  filter(RecordID %in% c(140501, 150649, 140936, 143656, 141264, 145611, 142998, 147514, 142731,150309, 155655, 156254)) %>%
  select(RecordID, Time)
```
```{r}
#FIND those RecordID that has only one entry
data %>%
  group_by(RecordID) %>%
  summarise(n = n()) %>%
  filter(n == 1)

```

```{r}
##Find all those lines with negative values (there cannot be according to the website)
data %>%
  filter(Albumin < 0 | ALP < 0 | ALT < 0 | AST < 0 | Bilirubin < 0 | BUN < 0 | Cholesterol < 0 | Creatinine < 0 | DiasABP < 0 | FiO2 < 0 | GCS < 0 | Glucose < 0 | HCO3 < 0 | HCT < 0 | HR < 0 | K < 0 | Lactate < 0 | Mg < 0 | MAP < 0 | Na < 0 | NIDiasABP < 0 | NIMAP < 0 | NISysABP < 0 | SysABP < 0 | PaCO2 < 0 | PaO2 < 0 | pH < 0 | Platelets < 0 | RespRate < 0 | SaO2 < 0 | Temp < 0 | TroponinI < 0 | TroponinT < 0 | Urine < 0 | WBC < 0)

#There arent anymore
```




# MULTIVARIATE GRAPHS
Consider also patients by demographic information such as gender, age (you can build bins
of e.g. 5 or 10 years), or the ICUType. Do you observe differences in the distributions of
some of the values across the groups?


## Boxplot for Age by ICUType

```{r}
data_static %>%
  filter(Age != -1) %>%
  ggplot(aes(x = factor(ICUType), y = Age, fill = factor(ICUType))) +
  geom_boxplot() +
  labs(title = "Age by ICU Type", x = "ICU Type", y = "Age") +
  theme_minimal()

# Boxplot for Weight by ICUType
data_static %>%
  filter(Weight != -1) %>%
  ggplot(aes(x = factor(ICUType), y = Weight, fill = factor(ICUType))) +
  geom_boxplot() +
  labs(title = "Weight by ICU Type", x = "ICU Type", y = "Weight") +
  theme_minimal()

# Boxplot for Height by ICUType
data_static %>%
  filter(Height != -1) %>%
  ggplot(aes(x = factor(ICUType), y = Height, fill = factor(ICUType))) +
  geom_boxplot() +
  labs(title = "Height by ICU Type", x = "ICU Type", y = "Height") +
  theme_minimal()
```

NOTE, I also tried to do some anova to see if the difference between ICU class is signficant. 
<!-- I get non signficant results for everything exept for Age, but then since the Age distribution is soo far from a Normal the results are not reliable. -->


### Chi-square ICUType vs MechVent
```{r}
data_chi <- data %>% filter(!is.na(ICUType) & !is.na(MechVent))
table_ICU_Mech <- table(data_chi$ICUType, data_chi$MechVent)
chisq.test(table_ICU_Mech) 
#check assumption 
chisq.test(table_ICU_Mech)$expected  #ok it's fine


library(lsr)
cramersV(table_ICU_Mech)

chisq.test(table_ICU_Mech)$stdres

mosaicplot(table_ICU_Mech, shade = TRUE, 
           main = "Mechanical Ventilation by ICU Type", 
           xlab = "ICU Type", ylab = "Mechanical Ventilation")  
```

```{r}
data_static %>% 
  filter(Age != -1) %>% 
  pull(Age) %>% 
  quantile(probs = seq(0, 1, 0.1)) %>%  # Change to seq(0,1,0.25) for quartiles
  knitr::kable(col.names = "Age")
```

### Gendere vs HR
I need to fill the Gender values for each person to having plots that resemble real data, not only the first row...
```{r}
data <- data %>%
  group_by(RecordID) %>%
  mutate(Gender = ifelse(any(Gender == 1, na.rm = TRUE), 
                         1,  # If any row has 1, set all to 1
                         0)) %>% # Otherwise, set all to 0
  ungroup()


data$Gender <- as.factor(data$Gender)  


data_hr <- data %>% filter(!is.na(HR))  # Only keep rows with HR

ggplot(data_hr, aes(x = as.factor(Gender), y = HR  , fill = as.factor(Gender))) +
  geom_violin(trim = FALSE) +
  labs(title = "Heart Rate by Gender", x = "Gender", y = "Heart Rate") +
  theme_minimal()

#perform t-test
t.test(HR ~ Gender, data = data_hr)
```

