---
title: "First EDA + Data Cleaning"
author: "Saverio Fontana"
date: "2025-03-16"
output: html_document
---

```{r setup, include=FALSE}
# Loading necessary libraries
library(arrow)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(DataExplorer)
knitr::opts_chunk$set(echo = TRUE)
```


```{r loading data}
setwd("..")
getwd()
# Read the Parquet file and load the data
data <- read_parquet("data/processed/set_a.parquet")
data$Time <- as.POSIXct(data$Time, tz = "UTC")  # Set time zone explicitly

# Check the first few timestamps
head(data$Time)  # Output should show the exact timestamps, e.g., "2025-03-10 01:00:00 UTC"

# Visualizing missing data
plot_missing(data)
```
## Initial Data Exploration

```{r}
# Check the dimensions of the dataset
dim(data)

# Check the column names
colnames(data)

# Summary statistics for numeric columns
summary(data)



# Histograms for numeric variables (ignoring NAs)
data %>%
  select(where(is.numeric)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Distributions of Numeric Variables (Ignoring NAs)")
```

### Variable Explanation and Notes

#### Variable Descriptions (general descriptors and time series)

"These six descriptors are collected at the time the patient is admitted to the ICU. 
Their associated time-stamps are 00:00 (thus they appear at the beginning of each 
patient's record)."

- **RecordID**: A unique integer for each ICU stay
- **Age**: Age in years
- **Gender**: Gender (0: female, 1: male)
- **Height**: Height in cm
- **ICUType**: ICU type (1: Coronary Care Unit, 2: Cardiac Surgery Recovery Unit, 3: Medical ICU, or 4: Surgical ICU)
- **Weight**: Weight in kg

"NB: All valid values for general descriptors, time series variables, and outcome-related 
descriptors are non-negative (≥ 0). A value of -1 indicates missing or unknown data."



```{r}
#So first of all we set the -1 in columns "Age", "Gender", "Height", ICUType, Weight to NA
# Replace -1 with NA for Height, Age, and Weight
data <- data %>%
  mutate(
    Height = ifelse(Height == -1, NA, Height),
    Age = ifelse(Age == -1, NA, Age),
    Weight = ifelse(Weight == -1, NA, Weight),
    ICUType = ifelse(ICUType == -1, NA, ICUType),
    Gender = ifelse(Gender ==-1, NA, Gender)
  )


#So when we will use algoritms that need to remove the NAs? What will we do? 
#I mean we could even do imputation actually (Weight can be predicted by age and Height for ex)
```


###Handling missing data in static variables
```{r}
# Static data: Keep the first row for each patient for simplicity in the next EDA
#and inspect for missing values
data_static <- data %>% group_by(RecordID) %>% slice(1)

# Frequency table for categorical variables (Gender, ICUType)
table(data_static$Gender, useNA = "ifany") 
table(data_static$ICUType, useNA = "ifany")

# Bar plots for categorical variables (Gender and ICUType)
data_static %>%
  ggplot(aes(x = factor(Gender))) +
  geom_bar() +
  labs(title = "Gender Distribution")

data_static %>%
  ggplot(aes(x = factor(ICUType))) +
  geom_bar() +
  labs(title = "ICU Type Distribution (Excluding NAs)")

```




### Exploring Continuous Static Variables (Height, Weight, Age)

```{r}
# Distribution for Height
data_static %>%
  filter(!is.na(Height)) %>%
  ggplot(aes(x = Height)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") + 
  labs(title = "Distribution of Height", x = "Height", y = "Count")


# Density plot for Height
data_static %>%
  filter(!is.na(Height)) %>%
  ggplot(aes(x = Height)) +
  geom_density(fill = "skyblue", alpha = 0.5) +  
  labs(title = "Density Plot of Height", x = "Height", y = "Density")

# Inspect unusually low or high Height values
outlier_data <- data_static %>%
  filter((Height < 100 | Height > 300) & Height != -1) %>%
  select(RecordID, Age, Height, Weight)

#upon inspection I'd suggest to set their values of Height to NA (maybe 1.8 for example can mean 180 cm but since we are not sure it's better to be conservative)

# Extract the RecordID from the filtered data
outlier_ids <- outlier_data$RecordID

# Set Height to NA for the identified records in the original data
data_clean <- data %>%
  mutate(Height = ifelse(RecordID %in% outlier_ids, NA, Height))

# Distribution for Weight
data_static %>%
  filter(Weight != -1) %>%
  ggplot(aes(x = Weight)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") + 
  labs(title = "Distribution of Weight", x = "Weight", y = "Count")

# Distribution for Age (NB, no values over 90, so maybe data are right truncated)
data_static %>%
  filter(Age != -1) %>%
  ggplot(aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") + 
  labs(title = "Distribution of Age", x = "Age", y = "Count")

# Smooth density plot for Weight
data_static %>%
  filter(Weight != -1) %>%
  ggplot(aes(x = Weight)) +
  geom_density(fill = "skyblue", alpha = 0.5) +  
  labs(title = "Density Plot of Weight", x = "Weight", y = "Density")

# Smooth density plot for Age
data_static %>%
  filter(Age != -1) %>%
  ggplot(aes(x = Age)) +
  geom_density(fill = "skyblue", alpha = 0.5) +  
  labs(title = "Density Plot of Age", x = "Age", y = "Density")

###########
# Inspecting Extreme Weight Values
###########

# Inspect unusually low or high Weight values
data_static %>%
  filter((Weight < 30 | Weight > 300) & Weight != -1) %>%  
  select(RecordID, Age, Height, Weight)
# Don't know actually, maybe it's an ok value honestly


#put also the summaries in the output
```
## Boxplot for Age by ICUType

```{r}
data_static %>%
  filter(Age != -1) %>%
  ggplot(aes(x = factor(ICUType), y = Age, fill = factor(ICUType))) +
  geom_boxplot() +
  labs(title = "Age by ICU Type", x = "ICU Type", y = "Age") +
  theme_minimal()

# Boxplot for Weight by ICUType
data_static %>%
  filter(Weight != -1) %>%
  ggplot(aes(x = factor(ICUType), y = Weight, fill = factor(ICUType))) +
  geom_boxplot() +
  labs(title = "Weight by ICU Type", x = "ICU Type", y = "Weight") +
  theme_minimal()

# Boxplot for Height by ICUType
data_static %>%
  filter(Height != -1) %>%
  ggplot(aes(x = factor(ICUType), y = Height, fill = factor(ICUType))) +
  geom_boxplot() +
  labs(title = "Height by ICU Type", x = "ICU Type", y = "Height") +
  theme_minimal()
```

NOTE, I also tried to do some anova to see if the difference between ICU class is signficant. 
<!-- I get non signficant results for everything exept for Age, but then since the Age distribution is soo far from a Normal the results are not reliable. -->




### Exploring Time changing variables

```{r}


```
```{r}

#Variables handling 

summary(data$Albumin) #(g/dL)
hist(data$Albumin)
"Albumin ok
bell shaped"

summary(data$ALP) #(IU/L)]
hist(data$ALP) #left skewed
#usually max range is around 150.
#Maybe should remove over 1000, Sticazzi


summary(data$ALT) #(IU/L)]
hist(data$ALT) #left skewed
#maybe should remove over 3000, sticazzi


summary(data$AST) #(IU/L)]
hist(data$AST) #left skewed
##maybe should remove over 3000, sticazzi

summary(data$Bilirubin) #(mg/dL)
hist(data$Bilirubin) #left skewed


summary(data$BUN) #(mg/dL)
hist(data$BUN) #left skewed
#GOOD VALUES


summary(data$Cholesterol) #(mg/dL)
hist(data$Cholesterol) #bell shaped
#okish values


summary(data$Creatinine) #(mg/dL)
hist(data$Creatinine) #left skewed
#okish 


summary(data$DiasABP) #(mmHg)
hist(data$DiasABP) #normali-ish
"Set to NaN if < 1	Unrealistic diastolic pressure
Set to NaN if > 200	Very high diastolic pressure, likely error"

summary(data$FiO2) #(0-1)
hist(data$FiO2) #impossible to define 


summary(data$GCS) #Glasgow Coma Scale (3-15) 
hist(data$GCS) #right skewed, makes sense, they are in coma ahahha

summary(data$Glucose) #(mg/dL)
hist(data$Glucose) #a bit normal (left skewed)

#find the columns RecordID with values over 1000
high_glucose_records <- data %>%
  filter(Glucose > 1000)

# Display the RecordID(s) with Glucose > 1000
IDs <- high_glucose_records$RecordID

#LET'S INSPECT One of them 
Unusual_patient <- data[data$RecordID == IDs[1], ]
Unusual_patient$Glucose
#An incredible drop in one hour. Don't know wheter we should cancel the over 1000



summary(data$HCO3) #(mmol/L)
hist(data$HCO3) #normal-ish


summary(data$HCT) #(%)
hist(data$HCT) #normal-ish

summary(data$HR) #(bpm)
hist(data$HR) #normal-ish
#There are some zeros, inspect:
high_HR_records <- data %>%
  filter(HR == 0)

#inspect one of them 
Unusual_patient <- data[data$RecordID == high_HR_records$RecordID[1], ]
Unusual_patient
#here, the 0 is most likely a mistake, or maybe a collapse and then a revitalization? 



Unusual_patient <- data[data$RecordID == high_HR_records$RecordID[3], ]
Unusual_patient
#but here ffor example no, cause it's the last recorded value and probably he is dead...



summary(data$K) #(mEq/L)
hist(data$K) #normal-ish-ish


summary(data$Lactate) #(mmol/L)
hist(data$Lactate) #left skewed
#some extreme values... 


summary(data$Mg) #(mmol/L)
hist(data$Mg) #normal-ish w some extreme values 


summary(data$MAP)
hist(data$MAP) #normal-ish
"MAP	Set to NaN if < 1	Impossible mean arterial pressure
Over 200?"

summary(data$MechVent) #1 if yes


summary(data$Na) #(mEq/L)
hist(data$Na) #normal-ish



summary(data$NIDiasABP) #(mmHg)
hist(data$NIDiasABP) #normal-ish


summary(data$NIMAP) #(mmHg)
hist(data$NIMAP) #normal-ish

summary(data$NISysABP) #(mmHg)
hist(data$NISysABP) #normal-ish





summary(data$SysABP) #(mmHg)
hist(data$SysABP) #normal-ish
"Set to NaN if < 1	Impossible systolic pressure
Over 200?"

summary(data$PaCO2) #(mmHg)
hist(data$PaCO2) #normal-ish, but the right tail is really big



summary(data$PaO2) #(mmHg)
hist(data$PaO2) #normal-ish, but the right tail is really big 
#inspect values smaller than 1
low_PaO2_records <- data %>%
  filter(PaO2 < 1)

# Display the RecordID(s) with PaO2 < 1
IDs <- low_PaO2_records$RecordID

#LET'S INSPECT One of them
Unusual_patient <- data[data$RecordID == IDs[1], ]
Unusual_patient
#the 0 seems more a missing value. 



#smaller than 20?
low_PaO2_records <- data %>%
  filter(PaO2 < 20)

# Display the RecordID(s) with PaO2 < 20
IDs <- low_PaO2_records$RecordID

#One of them is the one with 0, let's see the other
Unusual_patient <- data[data$RecordID == IDs[2], ]
Unusual_patient
#it's more likely to be 74.7


#Correct the non missing data changing 0 to Nan and changining the 74.7
data$PaO2[data$PaO2 == 0] <- NA
data$PaO2[data$PaO2 == 7.47] <- 74.7



summary(data$pH) #pH
# Multiply by 0.1 if between 65 and 80 (likely unit error)
idx <- which(!is.na(data$pH) & data$pH >= 65 & data$pH <= 80)
data$pH[idx] <- data$pH[idx] * 0.1

# Multiply by 0.01 if between 650 and 800 (likely unit error)
idx <- which(!is.na(data$pH) & data$pH >= 650 & data$pH <= 800)
data$pH[idx] <- data$pH[idx] * 0.01

hist(data$pH) #normal-ish


summary(data$Platelets) #(cells/nL)
hist(data$Platelets) #normal-ish ( a bit left skewed)



summary(data$RespRate) #(bpm)
hist(data$RespRate) #normal-ish

#same as HR, maybe mistakes, maybe dead, maybe ventilation
 



summary(data$SaO2) #(%)
hist(data$SaO2) #right skewed


summary(data$Temp) #(°C)
#probably mistakes in recording, also cause there are 17 patients w/ a value of -17.8 which we proceed to remove

data$Temp[!is.na(data$Temp) & data$Temp < 20] <- NA #under 20 is likely incompatible with life


summary(data$Temp)
hist(data$Temp) #normal-ish


summary(data$TroponinI) #(ng/mL)
hist(data$TroponinI) #left skewed
#some extreme values


summary(data$TroponinT) #(ng/mL)
hist(data$TroponinT) #left skewed
#some extreme values

summary(data$Urine) #(mL)
hist(data$Urine) #left skewed
#Some really extreme values

summary(data$WBC) #(cells/nL)
hist(data$WBC) #left skewed
#some extreme values


```

Save data back 
```{r}
# Save the cleaned data back to a Parquet file
getwd()
setwd("..")
write_parquet(data, "data/processed/set_a_cleaned.parquet")
```


These plots are from the data w/o forward fill to inspect the pure variables and their NAs. 
By visual and summary inspection many outliers are present. We handled some of them (lista di quello che abbiamo fatto), but for the handling of other extreme value a depp medical knowledge would be necessary, especially cause ICU patients can have values way more extreme than a normal patient. 
By visual insp